---
title: "Principles of inference in population ecology"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(WILD3810)
knitr::opts_chunk$set(echo = FALSE)
```


## Learning objectives: 

Ecological concepts:

> 1) What is the difference between a population and a sample? 

Programming concepts:

> 1) Creating objects in R  
> 2) Simulating random binomial samples


## Population

In a statistical sense, the **population** is the group of organisms that we wish to learn about. 

Note that the **statistical population** may or may not be the same as the **biological population**

The following code produces heights of 100 hypothetical individuals that make up our population:

```{r population, echo = TRUE}
N <- 100
mean_height <- 72 # mean height, inches
sd_height <- 5 # standard deviation of height

height <- rnorm(N, mean = mean_height, sd = sd_height)
```


```{r}
ggplot(data = data.frame(x = c(50, 90)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 72, sd = 5)) + ylab("") +
  scale_y_continuous(breaks = NULL)
```

## Parameters vs estimates

Estimates are usually differentiated from parameters by placing a hat over the symbol: $\hat{N}$. We will follow this convention through the course

## Sampling

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

