---
title: "Lecture 11"
subtitle: "Matrix population models"
author: "<br/><br/><br/>WILD3810 (Spring 2019)"
output:
  xaringan::moon_reader:
    css: ["default", "WILD3810.css", "WILD3810-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE)
library(WILD3810)
library(gganimate)
```

## Readings

---
## Assumptions of the B-D models

#### Over the coming weeks, we will learn about why and how to relax assumption 3:  


1) Population closed to immigration and emigration  
<br/>


2) Model pertains to only the limiting sex, usually females  
<br/>


**3) Birth and death rates are independent of an individual’s age or biological stage**  
<br/>


4) Birth and death rates are constant

--
### **How can we model population dynamics of populations with complex age structure?**

---
## Age-structured demography

### Consider the following life table:
<br/>

```{r}
library(kableExtra)
lt <- data.frame(x = c(0,1,2,3),
                       mx = c(0, "\\(m_1\\)", "\\(m_2\\)", "\\(m_3\\)"),
                       px = c("\\(P_0\\)", "\\(P_1\\)", "\\(P_2\\)", 0))

names(lt) <- c("\\(x\\)", "\\(m_x\\)", "\\(P_x\\)")

lt %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE, font_size = 22) 
```

???

What do $m_x$ and $P_x$ represent?  

What can you say about the reproductive and survival schedules of this population?

---
## Age-structred demography

#### List population abundance in discrete 1-year age classes $\LARGE (n_i)$  

- e.g., $\LARGE n_1$, $\LARGE n_2$, $\LARGE n_3$  

- $\LARGE n_i$ is the number of individuals about to experience their $\LARGE i^{th}$ birthday  

