---
title: "Lecture 8"
subtitle: "Life tables"
author: "<br/><br/><br/>WILD3810 (Spring 2019)"
output:
  xaringan::moon_reader:
    css: ["default", "WILD3810.css", "WILD3810-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE)
library(WILD3810)
library(gganimate)
```

## Readings

> Mills 84-92  

---
## Assumptions of the B-D models

#### Remember from the lecture 3 that our simple models of population growth were based on the following assumptions:  

--
1) Population closed to immigration and emigration  
<br/>

--
2) Model pertains to only the limiting sex, usually females  
<br/>

--
3) Birth and death rates are independent of an individual’s age or biological stage  
<br/>

--
4) Birth and death rates are constant


---
## Assumptions of the B-D models

#### In lectures 4 and 6, we learned about ways to model dynamics that do not conform to assumption 4:  


1) Population closed to immigration and emigration  
<br/>


2) Model pertains to only the limiting sex, usually females  
<br/>


3) Birth and death rates are independent of an individual’s age or biological stage  
<br/>


**4) Birth and death rates are constant**

???

In lecture 4, we learned about why birth and death rates might change as a function of abundance  

In lecture 6, we learned about ways that birth and death rates might vary stochastically  

---
## Assumptions of the B-D models

#### Over the coming weeks, we will learn about why and how to relax assumption 3:  


1) Population closed to immigration and emigration  
<br/>


2) Model pertains to only the limiting sex, usually females  
<br/>


**3) Birth and death rates are independent of an individual’s age or biological stage**  
<br/>


4) Birth and death rates are constant


---
class: inverse, center, middle

# Structured populations

---
## Age-structured populations

#### Survival and birth rates often vary with age

---
## Stage-structured populations

---
class: inverse, center, middle

# Age-structured populations

---
## Age-structured populations

#### **Life tables**  

- earliest accounting tool for calculating age-specific survival and mortality  


--
#### Applications to:  

-  Human demography  

- Insurance industry (actuarial sciences)  

- Health professions  


--
[Pearl and Parker (1921)](https://www.journals.uchicago.edu/doi/pdfplus/10.1086/279836) were the first to calculate a non-human life table  
- *Drosophila melanogaster* (a fruit fly)  

- Since used in ecology, evolution, and natural resource management  

---
## Life tables

#### Cohort life table

- follow group of individuals born within short period (a *cohort*) until each individual's death  

--
- develop life table for groups of individuals with similar traits (e.g., one for males, one for females)  

--
- **longitudinal** study design  

```{r fig.width=5, fig.height=3}
long <- data.frame(Time = seq(0, 100),
                   Age = seq(0, 100))

start <- data.frame(Time = c(0, 10, 20, 30, 40), 
                    Age = 0, 
                    Cohort = c("1", "2", "3", "4", "5"))

ggplot(long, aes(x = Time, y = Age)) +
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_continuous(limits = c(0, 100)) +
  geom_segment(aes(x = 0, xend = 95, y = 0, yend = 95), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "primary"]) +
  geom_segment(aes(x = 10, xend = 95, y = 0, yend = 85), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "warning"]) +
  geom_segment(aes(x = 20, xend = 95, y = 0, yend = 75), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "success"]) +
  geom_segment(aes(x = 30, xend = 95, y = 0, yend = 65), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "danger"]) +
  geom_segment(aes(x = 40, xend = 95, y = 0, yend = 55), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "info"]) +
  geom_point(data = start, aes(color = Cohort), size = 4) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

---
## Life tables

- $\Large x$: age of individuals

- $\Large N_x$: number of individuals observed alive at age $\Large x$ (sometimes called $\Large S_x$)
<br/>

```{r}

library(kableExtra)
lt <- data.frame(x = seq(from = 0, to  = 4),
                       Nx = c(100, 75, 50, 40, 0),
                       lx = "",
                       Lx = "",
                       Tx = "",
                       ex = "")

names(lt) <- c("\\(x\\)", "\\(N_x\\)", "\\(l_x\\)", "\\(L_x\\)", "\\(T_x\\)", "\\(e_x\\)")
```

```{r}
lt2 <- lt
names(lt2) <- c("\\(x\\)", "\\(N_x\\)", "", "", "", "")
lt2 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```


???

$S_x$ stands for number of *survivors* at age $x$

---
## Life tables

- $\Large l_x$: Probability of surviving to age $\Large x$

- $\Large l_x = N_x/N_0$
<br/>

```{r}
lt3 <- lt

names(lt3) <- c("\\(x\\)", "\\(N_x\\)", "\\(l_x\\)", "", "", "")
lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large l_x$: Probability of surviving to age $\Large x$

- $\Large l_x = N_x/N_0$
<br/>

```{r}
lt3[,3] <- c(1, "", "", "", "")
lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large l_x$: Probability of surviving to age $\Large x$

- $\Large l_x = N_x/N_0$
<br/>

```{r}
lt3[,3] <- c(1, 0.7, "", "", "")
lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large l_x$: Probability of surviving to age $\Large x$

- $\Large l_x = N_x/N_0$
<br/>

```{r}
lt3[,3] <- c(1, 0.7, 0.6, "", "")
lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large l_x$: Probability of surviving to age $\Large x$

- $\Large l_x = N_x/N_0$
<br/>

```{r}
lt3[,3] <- c(1, 0.7, 0.6, 0.4, "")
lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large l_x$: Probability of surviving to age $\Large x$

- $\Large l_x = N_x/N_0$
<br/>

```{r}
lt3[,3] <- c(1, 0.7, 0.6, 0.4, 0)
lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large P_x$: Probability of surviving **from** age $x$ **to** the next age (of those alive at age $x$)


- $\Large P_x = l_{x+1}/l_x$
<br/>

```{r}
lt4 <- lt3

names(lt4) <- c("\\(x\\)", "\\(N_x\\)", "\\(l_x\\)", "\\(P_x\\)", "", "")

lt4[,4] <- c(0.7, "", "", "", "")
lt4 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large P_x$: Probability of surviving **from** age $x$ **to** the next age (of those alive at age $x$)


- $\Large P_x = l_{x+1}/l_x$
<br/>

```{r}
lt4[,4] <- c(0.7, 0.857, "", "", "")
lt4 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large P_x$: Probability of surviving **from** age $x$ **to** the next age (of those alive at age $x$)


- $\Large P_x = l_{x+1}/l_x$
<br/>

```{r}

lt4[,4] <- c(0.7, 0.857, 0.667, "", "")
lt4 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large P_x$: Probability of surviving **from** age $x$ **to** the next age (of those alive at age $x$)


- $\Large P_x = l_{x+1}/l_x$
<br/>

```{r}

lt4[,4] <- c(0.7, 0.857, 0.667, 0, "-")
lt4 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large q_x$: Probability of dying **between** age $x$ the next age (of those alive at age $x$)


- $\Large q_x = 1 - P_x$
<br/>

```{r}
lt5 <- lt4

names(lt5) <- c("\\(x\\)", "\\(N_x\\)", "\\(l_x\\)", "\\(P_x\\)", "\\(q_x\\)", "")

lt5[,5] <- c(0.3, "", "", "", "")
lt5 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large q_x$: Probability of dying **between** age $x$ the next age (of those alive at age $x$)


- $\Large q_x = 1 - P_x$
<br/>

```{r}
lt5[,5] <- c(0.3, 0.143, "", "", "")
lt5 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large q_x$: Probability of dying **between** age $x$ the next age (of those alive at age $x$)


- $\Large q_x = 1 - P_x$
<br/>

```{r}

lt5[,5] <- c(0.3, 0.143, 0.333, "", "")
lt5 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

- $\Large q_x$: Probability of dying **between** age $x$ the next age (of those alive at age $x$)


- $\Large q_x = 1 - P_x$
<br/>

```{r}

lt5[,5] <- c(0.3, 0.143, 0.333, 1, "")
lt5 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```


---
## Life tables

- $\Large H_x$: Hazard (i.e., risk)  of dying **between** age $x$ the next age (of those alive at age $x$)


- $\Large H_x = -log(P_x)$
<br/>

```{r}
lt6 <- lt5

names(lt6) <- c("\\(x\\)", "\\(N_x\\)", "\\(l_x\\)", "\\(P_x\\)", "\\(q_x\\)", "\\(H_x\\)")

lt6[,6] <- c(0.357, 0.154, 0.405, "Inf", "")
lt6 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Life tables

#### Cohort tables

- Assume that all live individuals are available for observation at every age until death  


- Most readily applied to:  

    + Human populations (e.g., CDC NCHS)  
    + Plants  
    + Sessile animals  
    + Mobile animals on small islands with high observer detection  
    + Animals in captivity (zoos)  

---
## Life tables

#### Static tables

- assess number of individuals of known age at one point in time or the age of individuals dying at any point in time

```{r fig.width=7, fig.height=5}
long <- data.frame(Time = seq(0, 100),
                   Age = seq(0, 100))

start <- data.frame(Time = c(0, 10, 20, 30, 40), 
                    Age = 0, 
                    Cohort = c("1", "2", "3", "4", "5"))

ggplot(long, aes(x = Time, y = Age)) +
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_continuous(limits = c(0, 100)) +
  geom_rect(aes(xmin = 55, xmax = 65, ymin = -Inf, ymax = Inf), fill = "grey40", alpha = 0.9) + 
  geom_segment(aes(x = 0, xend = 95, y = 0, yend = 95), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "primary"]) +
  geom_segment(aes(x = 10, xend = 95, y = 0, yend = 85), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "warning"]) +
  geom_segment(aes(x = 20, xend = 95, y = 0, yend = 75), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "success"]) +
  geom_segment(aes(x = 30, xend = 95, y = 0, yend = 65), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "danger"]) +
  geom_segment(aes(x = 40, xend = 95, y = 0, yend = 55), size = 2, arrow = arrow(length = unit(0.03, "npc")),
               color = WILD3810_colors$value[WILD3810_colors$name == "info"]) +
  geom_point(data = start, aes(color = Cohort), size = 4) +
    geom_rect(aes(xmin = 55, xmax = 65, ymin = -Inf, ymax = Inf), fill = "grey80", alpha = 0.01) + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) 
```

---
## Static life tables

- $\Large a_x$: Deaths between age $\large x-1$ and age $\Large x$
<br/>

```{r}
lt3 <- lt

names(lt3) <- c("\\(x\\)", "\\(a_x\\)", "\\(N_x\\)", "", "", "")
lt3[,2] <- c("", 25, 15, 10, 7)
lt3[,3] <- c(57, "57-25=32", "32-15=17", "17-10=7", "7-7=0")

lt3 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```

---
## Static life tables

- $\Large a_x$: Deaths between age $\large x-1$ and age $\Large x$
<br/>

```{r}
lt4 <- lt3

names(lt4) <- c("\\(x\\)", "\\(a_x\\)", "\\(N_x\\)", "\\(l_x\\)", "\\(P_x\\)", "etc.")
lt4[,4] <- c(1, 0.56, 0.30, 0.12, 0)
lt4[,5] <- c(0.56, 0.54, 0.4, 0, "-")
lt4 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE) 
```


---
## Static life tables

#### Static life tables most readily applied to:

- Plants and animals where age can be accurately determined for all individuals at a given moment in time  

- e.g., using the rings/layering in trees, teeth, otoliths, horns, and maybe even telomere length on genes


---
## Static life tables

#### Static life table assumptions:

- All individuals have equivalent availability for observation, regardless of age  

- Stable age distribution  

- Survival does not change over time  

    + If it does, time effects may appear as ‘age effects’ and lead to biased results

---
class: inverse, middle, center

# Survivorship curves

---
## Survivorship curves

```{r fig.height=5 ,fig.width=7}
age <- seq(from = 0, to = 10, length.out = 100)
surv <- -1.2*age
lx <- exp(surv)

survII <- data.frame(Age = age, lx = lx, surv = surv)

p1 <- ggplot(survII, aes(Age, lx)) + geom_line() +
  scale_y_continuous(expression(l[x]))
p2 <- ggplot(survII, aes(Age, surv)) + geom_line() +
  scale_y_continuous(paste0("log(", expression(l[x]), ") 'Survivorship'"))

cowplot::plot_grid(p1, p2)

```

--
#### **Type II Survivorship curve** (Pearl 1928)
