---
title: "Gentle introduction to R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## What is R?

R is 

## Why R?

R is free(!), runs on pretty much every operating system, and has a huge user base. 

R is far from the only programming language that can be used to accomplish these tasks. But it is the most widely used language in the fields of ecology, evolution, and wildlife sciences. If you plan to pursue a career in any of these fields, proficiency in R is quickly becoming a prerequisite for many jobs. 

Even if you don't pursue a career in one of these fields, the ability to manipulate, analyze, and visualize data (otherwise known as *data science*) is an extremely marketable skill in many professions right now. 

## Where to get help


## My philosophy on learning R


## Using R - the very basics 

As a statistical programming tool, one thing R is very good at is doing math. So as a starting point, let's treat R like a fancy calculator. 

We interact with this calculator by typing number and operators (+, -, *, /) into the `Console` window.

Let's try it - in the window below (the Console), write the R code required to add two plus two and then click `Run Code`:

```{r two-plus-two, exercise=TRUE}

```

When you run the code, you should see the answer printed below the window. Play with your code a bit - try changing the number and the operators and then run the code again. 

## Creating objects

We can run R like a calculator by typing equations directly into the console and then printing the answer. But usually we don't want to just do a calculation and see the answer. Instead, we assign *values* to *objects*. That object is then saved in R's memory which allows us to use that object later in our analysis. 

This probably seems a bit confusing so let's try it. The following code creates an *object* called `x` and *assigns* it a value of `3`: 

```{r assignx, exercise=TRUE}
x <- 3
```

The operator `<-` is how we do assignments in R. Whatever is to the left of `<-` is the object's name and whatever is to the right is the value. As we will see later, objects can be much more complex than simply a number but for now, we'll keep it simple. 

**You try it - change the code above to create an object called `new.x`. Instead of assigning `new.x` a number, give it a calculation, for example `25/5`. What do you think the value of `new.x` is?**

## Working with objects

In the exercise above, you may have noticed that after running the code, R did not print anything. That is because we simply told R to create the object. However, now that it is stored in R's memory, we can do a lot of things with it. For one, we can print it to see the value. To do that, we simply type the name of the object and run the code:

```{r printx, exercise=TRUE}
new.x <- 25/5
new.x
```

We can also use objects to create new objects. *What do you think the following code does?*

```{r multx, exercise=TRUE}
x <- 3
y <- x*4
```

After running it, print the new object `y` to see its value. Were you right?

## Naming objects

It's a good idea to give objects names that tell you something about what the object represents. Names can be as long as you want them to be but should not have spaces (just remember long names require more typing so brevity is a good rule of thumb). Names also cannot start with a number and R is case-sensitive so, for example, `Apple` is *not* the same as `apple`. 

## Exercise 1: Lincoln-Peterson estimator

Remember from lecture 2 how we use the Lincoln-Peterson method to estimate the abundance of a closed population using mark-recapture data. As a reminder, the Lincoln-Peterson estimator is:

$$N = \frac{n_1n_2}{m_2}$$

where $n_1$ is the number of individuals captured (and marked) on the first sampling occasion, $n_2$ is the number of individuals captured on the second sampling occasion, and $m_2$ is the number of marked individuals in the second sample. 

Let's use R to estimate N for some made up data. Modify the following code to tell R that we captured 45 individuals on the first occasion, 32 on the second, and of those 32 sampled on the second occasion, 15 were marked:

```{r LP, exercise=TRUE}
n1 <-

n2 <-
  
m2 <- 
```

Now add, a new line of code below `m2 <- ` to estimate $N$ from our newly created objects. 

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

## Assignment

Each week
